---
title: "Interaction analysis in ADNI"
author: "Janou Roubroeks"
output: 
 html_document:
  toc: true
  toc_float: false
  number_sections: false
---

 ```{css, echo=FALSE}
    body .main-container {
      max-width: 1800px !important;
      margin-left: auto !important;
      margin-right: auto !important;
    }
    body {
      max-width: 1800px !important;
      margin-left: auto !important;
      margin-right: auto !important;
    }
    ```
    
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

wd<-""
packages<-c("wateRmelon", "dplyr", "qqman","kableExtra", "bacon","car")
setwd(wd)
lapply(packages, library, character.only=T) 

# Create a folder for storing output plots
if(!dir.exists(paste0(wd,"Plots"))){
  dir.create(paste0(wd,"Plots"))
}

if(!dir.exists(paste0(wd,"Combp"))){
  dir.create(paste0(wd,"Combp"))
}
```

# Interaction analysis - model  

This analysis of sex, age, diagnosis, and the interactions between these variables in the ADNI dataset follows the model that was previously run on the AddNeuroMed data:

```{r model, eval=FALSE}
lm(methylation ~ sex + age + diagnosis + age * sex + diagnosis * sex + age * diagnosis)
```

Packages used in this analysis: `r packages`.  
Results from this analysis are stored in "`r wd`".  


```{r data, echo=FALSE}
load("") # Load DNA methylation data (normalised)
Anno<-read.csv("", skip=7)
data<-betas(data.dasen)

Anno<-Anno[match(rownames(data), Anno$Name),]
rownames(Anno) <-Anno$IlmnID
	
```

# Regression results {.tabset} 


```{r Regression, echo=FALSE, warning=FALSE, message=FALSE}
# Regress out covariates 

# Function
RegrOut <- function( row,CD8T, CD4T, NK, Bcell, Mono){
  
  residuals <- try (
    resid(lm( row ~ CD8T+ CD4T+ NK+ Bcell+ Mono ,na.action=na.exclude)),
    silent=TRUE
  )
  residuals
} 


# Set cores for parallel processing 
cores<-32
cl<- makeCluster(cores)

### Apply model
RegData  <- {
  CD8T <- as.numeric(pheno[ colnames(data), 'CD8T'])
  CD4T <- as.numeric(pheno[ colnames(data), 'CD4T'])
  NK <- as.numeric(pheno[ colnames(data), 'NK'])
  Bcell <- as.numeric(pheno[ colnames(data), 'Bcell'])
  Mono <- as.numeric(pheno[ colnames(data), 'Mono'])
  
  t( parApply(cl, data, 1, RegrOut,CD8T, CD4T, NK, Bcell, Mono))
}

stopCluster(cl)


save(RegData, file="Regressed out data.RData")


# ANOVA

model <- function( row,Sex,Age, Diagnosis ){
  
  fit <- try (
    lm( row ~ as.factor(Sex) + Age + as.factor(Diagnosis) + as.factor(Sex)*Age + as.factor(Sex)*as.factor(Diagnosis) + Age*as.factor(Diagnosis)),
    silent=TRUE
  )
  results<-car::Anova(fit, type=3)
  as.numeric(as.matrix(results[,c(3:4)]))
  
 } 

# Set cores for parallel processing 
cores<-32
cl<- makeCluster(cores)

### Apply model

results  <- {
  Diagnosis   <- as.factor(pheno[ colnames(RegData), 'Diagnosis'  ])
  Sex   <- as.factor(pheno[ colnames(RegData), 'Sex'  ])
  Age   <- as.numeric(pheno[ colnames(RegData), 'Age'  ])
 
  
  t( parApply(cl,RegData, 1, model, Sex,Age,  Diagnosis ))
}

stopCluster(cl)
fit<-lm(RegData[1, ] ~ as.factor(Sex) + Age + as.factor(Diagnosis) + 
    as.factor(Sex) * Age + as.factor(Sex) * as.factor(Diagnosis) + 
    Age * as.factor(Diagnosis))
names<-rownames(car::Anova(fit,type=3))


names<-c(paste0(names, ".F"), paste0(names, ".p"))
names<-gsub("as.factor", '', names)
names<-gsub("\\(", '', names)
names<-gsub("\\)", '', names)
colnames(results)<-names

### Add annotation to model

rownames(Anno)<- Anno$IlmnID
Anno <-Anno[rownames(results), ]
merge(results, Anno, by="row.names") -> lm_Anno
rownames(lm_Anno)<- lm_Anno$Row.names


# save results
save(lm_Anno,results, file="Results Interaction.RData")

######## Create tables for publication #
lm_Anno$Position<- paste0("chr ",lm_Anno$CHR,": ", lm_Anno$MAPINFO)
lmRed<-lm_Anno[,c(1,grep("Position", colnames(lm_Anno)),
                  grep("Age|Sex", colnames(lm_Anno)),
                  grep("UCSC_RefGene_Name$|UCSC_RefGene_Group$|Relation_to_UCSC_CpG_Island$", colnames(lm_Anno)))]
colnames(lmRed)[c(1,c(ncol(lmRed)-2): ncol(lmRed))]<-c("ProbeID",
                   "UCSC Gene", "UCSC Gene Group","UCSC Gene Relation to CpG Island")

colnames(lmRed)<-gsub(":", ".", colnames(lmRed))


write.csv(lmRed[order(lmRed$Sex.p)[1:1000],], "Top1000 Sex.csv")
write.csv(lmRed[order(lmRed$Age.p)[1:1000],], "Top1000 Age.csv")
write.csv(lmRed[order(lmRed$`Sex.Age.p`)[1:1000],], "Top1000 SexAge.csv")
write.csv(lmRed[order(lmRed$`Sex.Diagnosis.p`)[1:1000],], "Top1000 SexDiag.csv")
write.csv(lmRed[order(lmRed$`Age.Diagnosis.p`)[1:1000],], "Top1000 AgeDiag.csv")

######## Create files for comb-p ########

DMRinput <- function(results,  chr = "CHR", 
                     mapinfo = "MAPINFO", p = "P"){
  dmr <- data.frame("chrom" = paste0("chr", as.character(results[ , chr])),
                    "start" = results[, mapinfo],
                    "end" = results[, mapinfo] + 1,
                    "pvalue" = results[ , p])
  dmr <- summarise(group_by(dmr, chrom, start), end = end, pvalue = pvalue)
  colnames(dmr) <- c("#chrom", "start", "end", "pvalue")
  return(dmr)
}

sex<-DMRinput(lm_Anno, p="Sex.p")
age<-DMRinput(lm_Anno, p="Age.p")
sexage<-DMRinput(lm_Anno, p="Sex:Age.p")
sexdiag<-DMRinput(lm_Anno, p="Sex:Diagnosis.p")
agediag<-DMRinput(lm_Anno, p="Age:Diagnosis.p")

write.table(sex, file = "./Combp/Sex_DMRinput.bed", sep = "\t", row.names = FALSE, col.names = T, quote = FALSE)
write.table(age, file = "./Combp/Age_DMRinput.bed", sep = "\t", row.names = FALSE, col.names = T, quote = FALSE)
write.table(sexage, file = "./Combp/SexAge_DMRinput.bed", sep = "\t", row.names = FALSE, col.names = T, quote = FALSE)
write.table(sexdiag, file = "./Combp/SexDiag_DMRinput.bed", sep = "\t", row.names = FALSE, col.names = T, quote = FALSE)
write.table(agediag, file = "./Combp/AgeDiag_DMRinput.bed", sep = "\t", row.names = FALSE, col.names = T, quote = FALSE)

# Prepare data for making tables below: 
sa<-grep("Sex.A", colnames(lmRed))
sd1<-grep("Sex.Diagnosis", colnames(lmRed))

ad1<-grep("Age.Diagnosis", colnames(lmRed))

int<-c(sa,sd1,ad1)
sex<-grep("Sex", colnames(lmRed))[!grep("Sex", colnames(lmRed)) %in% int]
age<-grep("Age", colnames(lmRed))[!grep("Age", colnames(lmRed)) %in% int]
anno<-grep("UCSC Gene$|UCSC Gene Group$|UCSC Gene Relation to CpG Island$",colnames(lmRed))
all<-c(sex,age,int)
st<-grep("ProbeID$|Position", colnames(lmRed))
```

## Sex:

Number of probes significantly associated with sex, passing the experiment wide threshold (2.4x10-7): `r nrow(lmRed[lmRed$Sex.p<2.4E-7,])`.  
*The top 50 sex associated DMPs:*  
  
```{r Top10sex, echo=FALSE}



lmRed<- lmRed[order(lmRed$Sex.p),]
select<-lmRed[1:50,c(st,sex,anno,all[!all %in% sex])]
rownames(select)<-NULL

for(c in grep("\\.", colnames(select))){
     for(i in 1:nrow(select)){
         if(abs(as.numeric(select[i,c]))>0.001){
             select[i,c]<-as.numeric(round(as.numeric(select[i,c]),3))
         } else {
             select[i,c]<- formatC(as.numeric(select[i,c]),format="e",digits=2)
         }}}

kable(select) %>%
  kable_styling(fixed_thead = T, bootstrap_options = c("striped"," hover")) %>%
scroll_box(width = "1000px", height = "500px")
```

## Age:  
Number of probes significantly associated with age, passing the experiment wide threshold (2.4x10-7): `r nrow(lmRed[lmRed$Age.p <2.4E-7,])`.  
  
*The top 50 age associated DMPs:*

```{r Top10age, echo=FALSE}

lmRed<- lmRed[order(lmRed$Age.p),]

select<-lmRed[1:50,c(st,age,anno,all[!all %in% age])]
rownames(select)<-NULL

for(c in grep("\\.", colnames(select))){
     for(i in 1:nrow(select)){
         if(abs(as.numeric(select[i,c]))>0.001){
             select[i,c]<-as.numeric(round(as.numeric(select[i,c]),3))
         } else {
             select[i,c]<- formatC(as.numeric(select[i,c]),format="e",digits=2)
         }}}

kable(select) %>%
  kable_styling(fixed_thead = T, bootstrap_options = c("striped"," hover")) %>%
scroll_box(width = "1000px", height = "500px")
```

## Interaction terms: {.tabset}

### Sex x Age:

Number of probes showing a significant interaction effect of sex and age, passing the experiment wide threshold (2.4x10-7): `r nrow(lmRed[lmRed$Sex.Age.p<2.4E-7,])`.

*The top 50 sex x age associated DMPs:*    

```{r Top10sexage1, echo=FALSE}

lmRed<- lmRed[order(lmRed$Sex.Age.p),]

select<-lmRed[1:50,c(st,sa,anno,all[!all %in% sa])]
rownames(select)<-NULL


for(c in grep("\\.", colnames(select))){
     for(i in 1:nrow(select)){
         if(abs(as.numeric(select[i,c]))>0.001){
             select[i,c]<-as.numeric(round(as.numeric(select[i,c]),3))
         } else {
             select[i,c]<- formatC(as.numeric(select[i,c]),format="e",digits=2)
         }}}

kable(select) %>%
  kable_styling(fixed_thead = T, bootstrap_options = c("striped"," hover")) %>%
scroll_box(width = "1000px", height = "500px")
```
  
### Sex x Diagnosis

Number of probes showing a significant interaction effect of sex and diagnosis, passing the experiment wide threshold (2.4x10-7): `r nrow(lmRed[lmRed$Sex.Diagnosis.p<2.4E-7,])`.

*The top 50 sex x diagnosis associated DMPs:*    

```{r Top10sexdiag1, echo=FALSE}
type<-sd1
lmRed<- lmRed[order(lmRed$Sex.Diagnosis.p),]


select<-lmRed[1:50,c(st,type,anno,all[!all %in% type])]
rownames(select)<-NULL


for(c in grep("\\.", colnames(select))){
     for(i in 1:nrow(select)){
         if(abs(as.numeric(select[i,c]))>0.001){
             select[i,c]<-as.numeric(round(as.numeric(select[i,c]),3))
         } else {
             select[i,c]<- formatC(as.numeric(select[i,c]),format="e",digits=2)
         }}}

kable(select) %>%
  kable_styling(fixed_thead = T, bootstrap_options = c("striped"," hover")) %>%
scroll_box(width = "1000px", height = "500px")
```


### Age x Diagnosis

Number of probes showing a significant interaction effect of age and diagnosis, passing the experiment wide threshold (2.4x10-7): `r nrow(lmRed[lmRed$Age.Diagnosis.p<2.4E-7,])`. 

*The top 50 age x diagnosis associated DMPs:*    

```{r Top10agediag1, echo=FALSE}
type<-ad1
lmRed<- lmRed[order(lmRed$Age.Diagnosis.p),]


select<-lmRed[1:50,c(st,type,anno,all[!all %in% type])]
rownames(select)<-NULL


for(c in grep("\\.", colnames(select))){
     for(i in 1:nrow(select)){
         if(abs(as.numeric(select[i,c]))>0.001){
             select[i,c]<-as.numeric(round(as.numeric(select[i,c]),3))
         } else {
             select[i,c]<- formatC(as.numeric(select[i,c]),format="e",digits=2)
         }}}

kable(select) %>%
  kable_styling(fixed_thead = T, bootstrap_options = c("striped"," hover")) %>%
scroll_box(width = "1000px", height = "500px")
```



# Inflation check {.tabset}

As both sex and age have large effects on DNA methylation, there may be some inflation of p-values. This can be inspected in QQ-plots of p-values and calculated lambda values as a measure of inflation.  

```{r QQplots,echo=FALSE}
as.data.frame(results)->results
colnames(results)<-gsub(":", ".", colnames(results))

qq(results$Sex.p, main = expression(paste('Q-Q plot of Sex ',italic('p'), '-values')),
   sub = paste0("Lambda: ", round( median(qchisq(1-results$Sex.p,1))/qchisq(0.5,1), 3)))

pdf("Plots/QQplots main Sex.pdf")
qq(results$Sex.p, main = expression(paste('Q-Q plot of Sex ',italic('p'), '-values')),
   sub = paste0("Lambda: ", round( median(qchisq(1-results$Sex.p,1))/qchisq(0.5,1), 3)))
invisible(dev.off())

qq(results$Age.p, main = expression(paste('Q-Q plot of Age ',italic('p'), '-values')),
   sub = paste0("Lambda: ", round( median(qchisq(1-results$Age.p,1))/qchisq(0.5,1), 3)))
pdf("Plots/QQplots main Age.pdf")
qq(results$Age.p, main = expression(paste('Q-Q plot of Age ',italic('p'), '-values')),
   sub = paste0("Lambda: ", round( median(qchisq(1-results$Age.p,1))/qchisq(0.5,1), 3)))
invisible(dev.off())

qq(results$Sex.Age.p, main = expression(paste('Q-Q plot of Sex*Age ',italic('p'), '-values')),
   sub = paste0("Lambda: ", round( median(qchisq(1-results$Sex.Age.p,1))/qchisq(0.5,1), 3)))
pdf("Plots/QQplots int SexAge.pdf")
qq(results$Sex.Age.p, main = expression(paste('Q-Q plot of Sex*Age ',italic('p'), '-values')),
   sub = paste0("Lambda: ", round( median(qchisq(1-results$Sex.Age.p,1))/qchisq(0.5,1), 3)))
invisible(dev.off())

qq(results$Sex.Diagnosis.p, main = expression(paste('Q-Q plot of Sex*Diagnosis ',italic('p'), '-values')),
   sub = paste0("Lambda: ", round( median(qchisq(1-results$Sex.Diagnosis.p,1))/qchisq(0.5,1), 3)))

pdf("Plots/QQplots int SexDiag.pdf")
qq(results$Sex.Diagnosis.p, main = expression(paste('Q-Q plot of Sex*Diagnosis ',italic('p'), '-values')),
   sub = paste0("Lambda: ", round( median(qchisq(1-results$Sex.Diagnosis.p,1))/qchisq(0.5,1), 3)))
invisible(dev.off())


qq(results$Age.Diagnosis.p, main = expression(paste('Q-Q plot of Age*Diagnosis ',italic('p'), '-values')),
   sub = paste0("Lambda: ", round( median(qchisq(1-results$Age.Diagnosis.p,1))/qchisq(0.5,1), 3)))
pdf("Plots/QQplots int AgeDiag.pdf")
qq(results$Age.Diagnosis.p, main = expression(paste('Q-Q plot of Age*Diagnosis ',italic('p'), '-values')),
   sub = paste0("Lambda: ", round( median(qchisq(1-results$Age.Diagnosis.p,1))/qchisq(0.5,1), 3)))
invisible(dev.off())
```
