---
title: "WGCNA of age, sex, and diagnosis interactions"
author: "Janou Roubroeks"
date:  "`r format(Sys.time(), '%d %B %Y')`"
output: 
 html_document:
  toc: true
  toc_float: false
  number_sections: false
---

 ```{css, echo=FALSE}
    body .main-container {
      max-width: 1800px !important;
      margin-left: auto !important;
      margin-right: auto !important;
    }
    body {
      max-width: 1800px !important;
      margin-left: auto !important;
      margin-right: auto !important;
    }
    ```

Overview of results for WGCNA associations of sex, age, sex x age, sex x diagnosis, and age x diagnosis to modules.
The modules were previously generated (also used in the Chapter 3 analysis of WGCNA for diagnosis and other traits).

An ANOVA interaction model was run on the module eigengenes as follows:


```{r model, eval=FALSE}
lm(MEs ~ sex + age + diagnosis + age * sex + diagnosis * sex + age * diagnosis)
```

Below is a heatmap showing the associations between the modules and traits of interest. The top value indicates the ANOVA F-statistic, and the p-value is shown in parentheses.

``` {r load, include=FALSE,echo=FALSE, warning=FALSE,message=FALSE}


setwd("")
packages<-c("wateRmelon", "WGCNA", "clusterProfiler", "KEGGREST", "kableExtra")
lapply(packages, library, character.only=T)
options(stringsAsFactors = FALSE)


######## Data preparation ########
# Data needed: WGCNA modules object, filtered methylation data used to create modules,
# full methylation data and phenotype file, data corrected for covariates, and 450K annotation file

load("/") # modules
load("") # filtered data (dat), used to create modules
load("") # full data
load("") # data corrected for covariates
Anno<-read.csv("")
Anno<-Anno[match(colnames(dat), Anno$Name),]

data.dasen.regr<-RegData
rm(RegData)

# Check that all files are in the correct order
# identical(rownames(pheno), rownames(dat))
# identical(colnames(data.dasen.regr), rownames(dat))

# Create a folder for storing output plots
if(!dir.exists(paste0(getwd(),"/Plots"))){
  dir.create(paste0(getwd(),"/Plots"))
}
if(!dir.exists(paste0(getwd(),"/Pathways"))){
  dir.create(paste0(getwd(),"/Pathways"))
}
```
  
``` {r plot, echo=FALSE, warning=FALSE,message=FALSE}
#### 1. Assign colours and calculate module eigengenes (MEs) ####

moduleColors = labels2colors(bwnet.fcx$colors)

MEs = moduleEigengenes(dat, moduleColors)$eigengenes
MEs = orderMEs(MEs)

# Disregard grey module from analysis, as it only contains probes that do not fit into any module
MEs<- MEs[,c(colnames(MEs)!="MEgrey")]


#### 2. Create Traits data ####

# Select only traits of interest (sex, age, and diagnosis in this case)
Traits<- pheno[ , c('Sex', "Age_At_Visit", "AtVisit_Group3")]
Traits$Sex<-as.character(Traits$Sex)
Traits$Sex<-as.numeric(Traits$Sex)

colnames(Traits)<-c("Sex", "Age", "Diagnosis")

#### 3. Regress out covariates #### 

# The established covariates still need to be regressed out of the ME data, as has been done for the methylation data
# Covariates to be regressed out: Cell type proportions and bisulfite batches (plate)

MET<-t(MEs)

# Function to regress out covariates
RegrOut <- function( row, CD8T, CD4T, NK, Bcell, Mono,
                     Plate1 , Plate2 , Plate3){
  
  residuals <- try (
    resid(lm( row ~ CD8T+ CD4T+ NK+ Bcell+ Mono+ Plate1 + Plate2 + Plate3 ,na.action=na.exclude)),
    silent=TRUE
  )
  if(inherits(residuals,'try-error')) return(rep(NA,ncol(MET)))
  residuals
} 

# Apply model
RegData  <- {
  CD8T <- as.numeric(pheno[ colnames(MET), 'CD8T'])
  CD4T <- as.numeric(pheno[ colnames(MET), 'CD4T'])
  NK <- as.numeric(pheno[ colnames(MET), 'NK'])
  Bcell <- as.numeric(pheno[ colnames(MET), 'Bcell'])
  Mono <- as.numeric(pheno[ colnames(MET), 'Mono'])
  Plate1 <-as.numeric(pheno[ colnames(MET), 'Plate1'])
  Plate2 <-as.numeric(pheno[ colnames(MET), 'Plate2'])
  Plate3 <-as.numeric(pheno[ colnames(MET), 'Plate3'])
  
  t( apply(MET, 1, RegrOut, CD8T, CD4T, NK, Bcell, Mono,
           Plate1 , Plate2 , Plate3))
}

as.data.frame(t(RegData))->RegData.t


#### 4. Outlier inspection ####

# Inspect values within modules for possible outliers
# Any values >5*sd will be set to NA

# Add names of outliers and the relevant module to an overview table ####
Overview<-data.frame(matrix(nrow=0,ncol=2))
for (i in 1:ncol(MEs)){
  module=colnames(MEs)[i]
  
  RegData.t[RegData.t[,module] < mean(RegData.t[,module],na.rm=T)-5*sd(RegData.t[,module],na.rm=T),module]->tes
  if (length(tes)>0){ 
      Overview<-rbind(Overview,c(rownames(RegData.t)[RegData.t[,module] %in% tes], module))
    }
  RegData.t[RegData.t[,module] > mean(RegData.t[,module],na.rm=T)+5*sd(RegData.t[,module],na.rm=T),module]->tes
  if (length(tes)>0){ 
    Overview<-rbind(Overview,c(rownames(RegData.t)[RegData.t[,module] %in% tes], module))
  }
}

# Plot stripcharts for later inspection
stripcharts<-function(module){
  par(mfrow=c(1,1))
  stripchart(RegData.t[ ,module], method="jitter",main=paste0(module, " Module Regressed Data"))
  abline(v=c(mean(RegData.t[ ,module]) +2*sd(RegData.t[ ,module])), col="orange")
  abline(v=c(mean(RegData.t[ ,module]) -2*sd(RegData.t[ ,module])), col="orange")
  abline(v=c(mean(RegData.t[ ,module]) +5*sd(RegData.t[ ,module])), col="red")
  abline(v=c(mean(RegData.t[ ,module]) -5*sd(RegData.t[ ,module])), col="red")
  
}

 for (i in 1:ncol(RegData.t)){
   stripcharts(colnames(RegData.t)[i])
 }


# Only 2 outliers; set to NA in specific modules
RegData.t[Overview[1,1], Overview[1,2]]<-NA
RegData.t[Overview[2,1], Overview[2,2]]<-NA

# Plots following removal
 
 for (i in 1:ncol(RegData.t)){
   stripcharts(colnames(RegData.t)[i])
 }


#### 5. Module-Trait Correlations #####

# Run an ANOVA interaction model on the MEs, as done for the main analysis in Chapter 4

model <- function( row,Sex,Age, Diagnosis ){
  
  fit <- try (
    lm( row ~ as.factor(Sex) + Age + as.factor(Diagnosis) + as.factor(Sex)*Age + as.factor(Sex)*as.factor(Diagnosis) + Age*as.factor(Diagnosis)),
    silent=TRUE
  )
  results<-car::Anova(fit, type=3)
  as.numeric(as.matrix(results[,c(3:4)]))
  
} 

# Set cores for parallel processing 
cores<-18
cl<- makeCluster(cores)

### Apply model

results  <- {
  Diagnosis   <- as.factor(pheno[ colnames(RegData), 'AtVisit_Group3'  ])
  Sex   <- as.factor(pheno[ colnames(RegData), 'Sex'  ])
  Age   <- as.numeric(pheno[ colnames(RegData), 'Age_At_Visit'  ])
  
  
  t( parApply(cl,RegData, 1, model, Sex,Age,  Diagnosis ))
}

stopCluster(cl)
fit<-lm(RegData[1, ] ~ as.factor(Sex) + Age + as.factor(Diagnosis) + 
          as.factor(Sex) * Age + as.factor(Sex) * as.factor(Diagnosis) + 
          Age * as.factor(Diagnosis))
names<-rownames(car::Anova(fit,type=3))

names<-c(paste0(names, ".F"), paste0(names, ".p"))
names<-gsub("as.factor", '', names)
names<-gsub("\\(", '', names)
names<-gsub("\\)", '', names)
colnames(results)<-names

Cor<-results[,c(2:3,5:7)]
Cor.p<-results[,c(10:11,13:15)]
```
```{r heatmap, echo=FALSE, warning=FALSE,message=FALSE, fig.height=7, fig.width=7}
# Heatmap of correlations

textMatrix = paste(signif(t(Cor), 2), "\n(",
                   signif(t(Cor.p), 1), ")", sep = "");
dim(textMatrix) = dim(t(Cor))
par(mar = c(4,6,3,1));

colfunc <- colorRampPalette(c("white", "indianred3"))
labeledHeatmap(Matrix = Cor,
               xLabels = gsub(".F", "",colnames(Cor)),
               yLabels = colnames(RegData.t),
               ySymbols = colnames(RegData.t),
               colorLabels = T,
               colors = colfunc(6),
               textMatrix = t(textMatrix),
               setStdMargins = FALSE,
               cex.text = 0.7,
               cex.lab = 0.7,
               zlim = c(0,6),
               main = paste("Module-trait relationships"))

  
```    
  

Modules showing a significant association to age: greenyellow, pink, green  
Modules showing a significant association to sex x diagnosis: salmon   

MM to PS correlations were calculated for these modules and traits:  

```{r data, echo=FALSE, warning=FALSE,message=FALSE}

#### 6. Calculate and correlate MM/PS for modules associated to traits ####
# Assess whether probes central to a module, are also the ones driving the association with a trait

# MM: correlate probe methylation to regressed out MEs, = pearson
# PS: correlate probe methylation to trait of interest, = pearson for continuous, spearman for ordinal
# Then perform pearson correlations between MM and PS
Cor.p<-as.data.frame(Cor.p)



### 6.1 Load Functions ####

# MMPS plots function:
# the function verboseScatterplot (for MMPS plots) is supposed to always calculate the pearson correlations,
# but it actually shows spearman correlations. Have created a new function to create MMPS (full and top 15% MM) plots
source("MMPSplot_function.R")

# 1.2. MMPS correlations function
source("MMPScor_function.R")

### 6.2. Calculate Module Memberships ####

dat.original <- dat
data.dasen.regr <- t(data.dasen.regr)
dat <- data.dasen.regr[,colnames(dat.original)]

modNames = rownames(RegData)
probeModuleMembership = as.data.frame(cor(dat, RegData.t, method="pearson",use = "pairwise.complete.obs"))

# Wrote a new loop to calculate MM p-value, as original code corrected for total samples, not pairwise
 for(o in 1:ncol(RegData.t)){
   for (i in 1:ncol(dat)){
     x<-cor.test(RegData.t[,o], dat[,i],method="pearson",use="pairwise.complete.obs")$p.value
     if(i==1){
       x->Pval.MM
     } else{
       Pval.MM<-cbind(Pval.MM,x)
     }
   }
   if(o==1){
     MMPvalue<-Pval.MM
   }else{
     MMPvalue<-rbind(MMPvalue,Pval.MM)
   }
 }
 colnames(MMPvalue)<-colnames(dat)
 rownames(MMPvalue)<-colnames(RegData.t)
 t(MMPvalue)->MMPvalue


names(probeModuleMembership) = paste("MM", modNames, sep="");
names(MMPvalue) = paste("p.MM", modNames, sep="");


### 6.3. Probe Significance: Age ####

Trait = as.data.frame(pheno$Age_At_Visit);
names(Trait) = "Age"

probeTraitSignificance = as.data.frame(cor(dat, Trait, method="pearson", use = "pairwise.complete.obs"));
PSPvalue <-as.data.frame(apply(dat,2, function(r) cor.test(r,Trait[,1],method="pearson",
                                                           use="pairwise.complete.obs",exact=F)$p.value))
colnames(PSPvalue)<-colnames(Trait)
names(probeTraitSignificance) = paste("PS.", names(Trait), sep="");
names(PSPvalue) = paste("p.PS.", names(Trait), sep="");


# 3.2 Plot 

MMPSplot("greenyellow", "chartreuse3")
MMPSplot("pink", "deeppink2")
MMPSplot("green", "green3")

# 3.3 Correlate MM to PS

MMPScor("greenyellow")
MMPScor("pink")
MMPScor("green")

### 6.4. Probe Significance: Sex*Diagnosis ####

pheno$Sex<-as.character(pheno$Sex)
pheno$Sex<-as.numeric(pheno$Sex)
pheno$SexDiag<- c(pheno$Sex *pheno$AtVisit_Group3)
Trait = as.data.frame(pheno$SexDiag);
names(Trait) = "Sex*Diagnosis"


probeTraitSignificance = as.data.frame(cor(dat, Trait, method="spearman", use = "pairwise.complete.obs"));
PSPvalue <-as.data.frame(apply(dat,2, function(r) cor.test(r,Trait[,1],method="spearman",
                                                           use="pairwise.complete.obs",exact=F)$p.value))
colnames(PSPvalue)<-colnames(Trait)
names(probeTraitSignificance) = paste("PS.", names(Trait), sep="");
names(PSPvalue) = paste("p.PS.", names(Trait), sep="");


# 3.2 Plot 

MMPSplot("salmon")

# 3.3 Correlate MM to PS

MMPScor("salmon")

	
```
  
A table of MM to PS correlations:  
```{r MMPS, echo=FALSE}
kable(MMPScorrelation)%>%
  kable_styling(fixed_thead = T, bootstrap_options = c("striped"," hover"))
```  

# Pathway analyses:  
Pathway analyses have been run for the green, greenyellow, pink and salmon modules.
Plots are automatically generated below if there are any FDR-significant pathways.



```{r pathwaysGreen, echo=FALSE, warning=FALSE,message=FALSE}

t(dat)->dat
names(moduleColors)<-rownames(dat)


#### green all ####
colour<- "green"

# Specify project name to be appended to file names
projectName <- paste0(colour," module")

## Load required packages / functions
source("/mnt/data1/Janou/References/crystalmeth_fun.R")
read.csv(paste0("Pathways/", colour," GO Gene Analysis Output WGCNA_fullbg.csv"))->go
read.csv(paste0("Pathways/", colour," KEGG Gene Analysis Output WGCNA_fullbg.csv"))->kegg
# Plots

topplotGO<-function(Top, Title){
  Top$order<-c(1:nrow(Top))
  Top$qvalue<-as.numeric(scientific(Top$FDR))

  ggplot(Top, aes_string(x = reorder(Top$Term, -Top$order), y = Top$DE,
                       fill = Top$qvalue))  + scale_fill_continuous(low = "red", 
                                                                    high = "blue", name = "FDR-value", guide = guide_colorbar(reverse = TRUE))+
  geom_bar(stat="identity")+coord_flip()+xlab(NULL)+ylab(NULL)+theme_bw()+
  ggtitle(paste0("GO Terms in ", Title ))
}


topplotKEGG<-function(Top, Title){
Top$order<-c(1:nrow(Top))
Top$qvalue<-as.numeric(scientific(Top$FDR))

ggplot(Top, aes_string(x = reorder(Top$desc, -Top$order), y = Top$DE,
                       fill = Top$qvalue))  + scale_fill_continuous(low = "red", 
                                                                    high = "blue", name = "FDR-value", guide = guide_colorbar(reverse = TRUE))+
  geom_bar(stat="identity")+coord_flip()+xlab(NULL)+ylab(NULL)+theme_bw()+
  ggtitle(paste0("KEGG Terms in ",Title))

}

Title<-projectName
if(length(go[go$FDR<0.05,1])>0 & length(go[go$FDR<0.05,1])<=20){
    Top<-go[go$FDR<0.05,]
    topplotGO(Top,Title)
} else if(length(go[go$FDR<0.05,1])>20) {
  Top<-go[1:20,]
  topplotGO(Top,Title)
}

if(length(kegg[kegg$FDR<0.05,1])>0 & length(kegg[kegg$FDR<0.05,1])<=20){
  Top<-kegg[kegg$FDR<0.05,]
topplotKEGG(Top,Title)
} else if(length(kegg[kegg$FDR<0.05,1])>20) {
  Top<-kegg[1:20,]
  topplotKEGG(Top,Title)
}
``` 
  

Number of GO terms FDR-significant in the `r colour` module: `r sum(go$FDR < 0.05)`.  
Number of KEGG terms FDR-significant in the `r colour` module: `r sum(kegg$FDR < 0.05)`.  

  
```{r pathwaysGY, echo=FALSE, warning=FALSE,message=FALSE}
#### greenyellow all ####
colour<- "greenyellow"

# Specify project name to be appended to file names
projectName <- paste0(colour," module")

read.csv(paste0("Pathways/", colour," GO Gene Analysis Output WGCNA_fullbg.csv"))->go
read.csv(paste0("Pathways/", colour," KEGG Gene Analysis Output WGCNA_fullbg.csv"))->kegg

Title<-projectName
if(length(go[go$FDR<0.05,1])>0 & length(go[go$FDR<0.05,1])<=20){
    Top<-go[go$FDR<0.05,]
    topplotGO(Top,Title)
} else if(length(go[go$FDR<0.05,1])>20) {
  Top<-go[1:20,]
  topplotGO(Top,Title)
}

if(length(kegg[kegg$FDR<0.05,1])>0 & length(kegg[kegg$FDR<0.05,1])<=20){
  Top<-kegg[kegg$FDR<0.05,]
topplotKEGG(Top,Title)
} else if(length(kegg[kegg$FDR<0.05,1])>20) {
  Top<-kegg[1:20,]
  topplotKEGG(Top,Title)
}
```
  
Number of GO terms FDR-significant in the `r colour` module: `r sum(go$FDR < 0.05)`.  
Number of KEGG terms FDR-significant in the `r colour` module: `r sum(kegg$FDR < 0.05)`.  

```{r pathwaysPink, echo=FALSE, warning=FALSE,message=FALSE}
#### pink all ####
colour<- "pink"


# Specify project name to be appended to file names
projectName <- paste0(colour," module")

read.csv(paste0("Pathways/", colour," GO Gene Analysis Output WGCNA_fullbg.csv"))->go
read.csv(paste0("Pathways/", colour," KEGG Gene Analysis Output WGCNA_fullbg.csv"))->kegg


Title<-projectName
if(length(go[go$FDR<0.05,1])>0 & length(go[go$FDR<0.05,1])<=20){
    Top<-go[go$FDR<0.05,]
    topplotGO(Top,Title)
} else if(length(go[go$FDR<0.05,1])>20) {
  Top<-go[1:20,]
  topplotGO(Top,Title)
}

if(length(kegg[kegg$FDR<0.05,1])>0 & length(kegg[kegg$FDR<0.05,1])<=20){
  Top<-kegg[kegg$FDR<0.05,]
topplotKEGG(Top,Title)
} else if(length(kegg[kegg$FDR<0.05,1])>20) {
  Top<-kegg[1:20,]
  topplotKEGG(Top,Title)
}
```
   
Number of GO terms FDR-significant in the `r colour` module: `r sum(go$FDR < 0.05)`.  
Number of KEGG terms FDR-significant in the `r colour` module: `r sum(kegg$FDR < 0.05)`.  

```{r pathwaysSal, echo=FALSE, warning=FALSE,message=FALSE}
#### salmon all ####
colour<- "salmon"

# Specify project name to be appended to file names
projectName <- paste0(colour," module Sex*Diagnosis")


read.csv(paste0("Pathways/", colour," GO Gene Analysis Output WGCNA_fullbg.csv"))->go
read.csv(paste0("Pathways/", colour," KEGG Gene Analysis Output WGCNA_fullbg.csv"))->kegg

Title<-projectName
if(length(go[go$FDR<0.05,1])>0 & length(go[go$FDR<0.05,1])<=20){
    Top<-go[go$FDR<0.05,]
    topplotGO(Top,Title)
} else if(length(go[go$FDR<0.05,1])>20) {
  Top<-go[1:20,]
  topplotGO(Top,Title)
}

if(length(kegg[kegg$FDR<0.05,1])>0 & length(kegg[kegg$FDR<0.05,1])<=20){
  Top<-kegg[kegg$FDR<0.05,]
topplotKEGG(Top,Title)
} else if(length(kegg[kegg$FDR<0.05,1])>20) {
  Top<-kegg[1:20,]
  topplotKEGG(Top,Title)
}


```

   
Number of GO terms FDR-significant in the `r colour` module: `r sum(go$FDR < 0.05)`.  
Number of KEGG terms FDR-significant in the `r colour` module: `r sum(kegg$FDR < 0.05)`.  
